---
title: "NHANES 2017-2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Creation of the Directed acyclic graph (DAG)"
author: "Amanda Millatt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup-env, fig.width=10, fig.height=8}
# ---- packages ----
pkgs <- c("dagitty", "ggdag", "ggplot2", "dplyr")

to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

theme_set(theme_minimal(base_size = 12))
```

# Defining the relationships for the Directed acyclic graph (DAG)

The structural relationships between health insurance coverage, blood
sugar screening, and potential confounding pathways are represented in
the Directed Acyclic Graph (DAG) below. 

**Methods** To ensure conditional exchangeability, peer-reviewed literature was assessed to identify known predictors of both health insurance coverage and preventative diabetes screening. 

The DAG was first created using DAGGITY online and then modifed using the code below. 

**Potential Confounders** Below are the the covariates included in the DAG:

- *Demographic / socioeconomic:* age, sex, race/ethnicity, education, family
  income-to-poverty ratio, marital status, country of birth, interview
  language.
- *Health status / risk:* BMI, self-rated general health, prior hypertension
  diagnosis. 
- *Health behavior:* ever-smoking.

```{r daggity}
nhanes_dag <- dagitty('dag {
"Blood Sugar Screening (Y)" [outcome,pos="1.058,1.374"]
"Country of birth" [pos="-1.175,-1.600"]
"Family income-to-poverty ratio" [pos="1.128,-1.885"]
"Health-insurance coverage (A)" [exposure,pos="-1.588,1.323"]
"High blood pressure" [pos="-0.352,0.592"]
"Interview language" [pos="0.635,-2.433"]
"Marital Status" [pos="-1.823,-2.031"]
"Race/Ethnicity" [pos="-0.921,-0.964"]
"Self-rated general health" [pos="0.987,-0.446"]
"Smoking status" [pos="-1.856,-0.606"]
Acculturation [latent,pos="1.161,-2.389"]
Age [pos="-0.291,-1.622"]
BMI [pos="0.287,-1.359"]
Education [pos="-0.925,-2.382"]
Sex [pos="0.108,-0.241"]
"Country of birth" -> "Blood Sugar Screening (Y)"
"Country of birth" -> "Health-insurance coverage (A)"
"Country of birth" -> "Interview language"
"Family income-to-poverty ratio" -> "Blood Sugar Screening (Y)"
"Family income-to-poverty ratio" -> "Health-insurance coverage (A)"
"Family income-to-poverty ratio" -> "Self-rated general health"
"Health-insurance coverage (A)" -> "Blood Sugar Screening (Y)"
"High blood pressure" -> "Blood Sugar Screening (Y)"
"High blood pressure" -> "Health-insurance coverage (A)"
"Interview language" -> "Blood Sugar Screening (Y)"
"Interview language" -> "Health-insurance coverage (A)"
"Interview language" -> Education
"Marital Status" -> "Family income-to-poverty ratio"
"Marital Status" -> "Health-insurance coverage (A)"
"Marital Status" -> "High blood pressure"
"Race/Ethnicity" -> "Blood Sugar Screening (Y)"
"Race/Ethnicity" -> "Health-insurance coverage (A)"
"Race/Ethnicity" -> "High blood pressure"
"Self-rated general health" -> "Blood Sugar Screening (Y)"
"Self-rated general health" -> "Health-insurance coverage (A)"
"Smoking status" -> "Blood Sugar Screening (Y)"
"Smoking status" -> "Health-insurance coverage (A)"
"Smoking status" -> "High blood pressure"
Acculturation -> "Interview language"
Age -> "Blood Sugar Screening (Y)"
Age -> "Health-insurance coverage (A)"
Age -> "High blood pressure"
BMI -> "Blood Sugar Screening (Y)"
BMI -> "Health-insurance coverage (A)"
BMI -> "High blood pressure"
BMI -> "Self-rated general health"
Education -> "Blood Sugar Screening (Y)"
Education -> "Family income-to-poverty ratio"
Education -> "Health-insurance coverage (A)"
Education -> "High blood pressure"
Sex -> "Blood Sugar Screening (Y)"
Sex -> "Health-insurance coverage (A)"
Sex -> "Smoking status"
}')
```

# Generating the DAG a format defined by the ggdag package

To create a readable DAG, the ggdag package uses a mixure of r-code and html. 

The only unadjusted variable is *High blood pressure* due to it being a collider in multiple pathhways from the exposure to the outcome. 


```{r layout}
dag_cleaned <- tidy_dagitty(nhanes_dag) %>%
  node_status() %>% 
  mutate(
    x     = x * 1.5,
    y     = y * 1.5,
    xend  = xend * 1.5,
    yend  = yend * 1.5,
    
  
    status = case_when(
      name == "Health-insurance coverage (A)" ~ "Exposure",
      name == "Blood Sugar Screening (Y)" ~ "Outcome",
      name %in% c("Acculturation") ~ "Unmeasured", 
      name == "High blood pressure" ~ "Unadjusted",
      TRUE ~ "Adjusted" 
    )
  )

ggplot(dag_cleaned, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    arrow_directed = grid::arrow(length = unit(3, "mm"), type = "closed"),
    edge_colour = "black",
    edge_width = 0.5
  ) +

  geom_dag_node(aes(color = status), size = 12) +

  geom_dag_label_repel(
    aes(label = name),
    direction = "y",
    size = 2,
    box.padding = 0.1,
    point.padding = 0.1,
    segment.color = "grey75",
    segment.size = 0.5,
    label.size = 0.1,
    fill = "white",
    fontface = "bold",
    show.legend = FALSE
  ) +
  theme_dag() +
  scale_color_manual(
    values = c(
      "Exposure"   = "#1f77b4", 
      "Outcome"    = "#d62728", 
      "Adjusted"   = "#2ca02c",
      "Unadjusted" = "#ff7f0e",
      "Unmeasured" = "#7f7f7f"
    ),
    name = "Variable Status"
  ) +
  labs(
    title = "Directed Acyclic Graph (DAG) for 2017-2020 NHANES Data",
    subtitle = "Exposure: Insurance Status | Outcome: Blood Sugar Screening"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey30"),
    plot.margin = margin(15, 15, 15, 15)
  )
```