---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Exploratory covariate analysis and positivity check"
author: "Sujit Silas Armstrong"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("haven", "dplyr", "tidyr", "purrr", "forcats",
          "ggplot2", "gtsummary", "labelled", "scales")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

theme_set(theme_minimal(base_size = 12))
```


# 1. Causal estimand

We fix the causal question before any exploratory analysis so that covariate
selection and filtering decisions are driven by the assumptions, not by the data.
 
- **Target population.** Non-institutionalized U.S. adults aged ≥ 20 years
  sampled in the NHANES **2017–March 2020 pre-pandemic** cycle (the `P_` files).
- **Treatment / exposure $A$.** Health-insurance coverage,
  `HIQ011` — *"Are you covered by health insurance or some other kind of
  health-care plan?"* Binarized as **Insured (1)** vs. **Uninsured (0)**.
- **Outcome $Y$.** `DIQ180` — *"Had blood tested for sugar/diabetes in the past
  three years?"* Binarized as **Screened (1)** vs. **Not screened (0)**.
- **Estimand.** The **average treatment effect (ATE)** on the probability of
  blood-sugar screening:
  $$\tau \;=\; \mathbb{E}\big[\,Y^{a=1} - Y^{a=0}\,\big]
            \;=\; \mathbb{E}\big[\,Y^{\text{insured}} - Y^{\text{uninsured}}\,\big].$$

**Identifying assumptions.** The ATE is identified from observational NHANES
data under:

1. **Consistency** — observed $Y$ equals the potential outcome under the
   observed insurance status.
2. **Conditional exchangeability (no unmeasured confounding)** —
   $Y^a \perp A \mid L$ for the measured covariate set $L$ below.
3. **Positivity** — $0 < \Pr(A = a \mid L = l) < 1$ for every covariate
   stratum $l$ that occurs in the population. **This is the assumption the
   summary table in Section 6 is built to interrogate.**

**Measured confounders $L$.** Factors plausibly affecting *both* insurance
coverage and the propensity to be screened:

- *Demographic / socioeconomic:* age, sex, race/ethnicity, education, family
  income-to-poverty ratio, marital status, U.S.-vs-foreign birth, interview
  language (acculturation proxy).
- *Health status / risk:* BMI, self-rated general health, prior hypertension
  diagnosis (a guideline risk factor that itself prompts diabetes screening).
- *Health behavior:* ever-smoking, ever-drinking alcohol.

# 2. Download the NHANES source files

The 2017–2020 pre-pandemic files are served as SAS transport (`.XPT`) files from
the CDC. We download each once into a local `data/` cache and read with
`haven::read_xpt()`.

```{r download}
data_dir <- "data"
dir.create(data_dir, showWarnings = FALSE)

base_url <- "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/"

# component file -> the variables we keep from it
files <- c(
  P_DEMO = "P_DEMO.xpt",  # demographics: age, sex, race, education, income, lang
  P_HIQ  = "P_HIQ.xpt",   # health insurance questionnaire (treatment)
  P_DIQ  = "P_DIQ.xpt",   # diabetes questionnaire (outcome)
  P_BMX  = "P_BMX.xpt",   # body measures: BMI
  P_SMQ  = "P_SMQ.xpt",   # smoking: ever smoked >=100 cigarettes
  P_ALQ  = "P_ALQ.xpt",   # alcohol: ever had a drink
  P_HUQ  = "P_HUQ.xpt",   # health-care utilization: self-rated general health
  P_BPQ  = "P_BPQ.xpt"    # blood pressure: prior hypertension diagnosis
)

# imap() calls .f(value, name); `file` is the .xpt filename, `name` the label
read_nhanes <- function(file, name) {
  dest <- file.path(data_dir, file)
  if (!file.exists(dest)) {
    utils::download.file(paste0(base_url, file), dest, mode = "wb", quiet = TRUE)
  }
  haven::read_xpt(dest)
}

raw <- imap(files, read_nhanes)
map(raw, dim)
```

# 3. Merge on the respondent ID (`SEQN`)

`SEQN` is the unique respondent sequence number that links every NHANES file.
We keep all demographics records and left-join the questionnaire/exam files, so
we start with the complete sampled population and *nothing is dropped yet*.

```{r merge}
demo <- raw$P_DEMO %>%
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, DMDEDUC2, DMDMARTZ,
         DMDBORN4, INDFMPIR, SIALANG,
         # survey-design variables, carried for later weighted analysis
         SDMVPSU, SDMVSTRA, WTINTPRP, WTMECPRP)

hiq <- raw$P_HIQ %>% select(SEQN, HIQ011)
diq <- raw$P_DIQ %>% select(SEQN, DIQ180)
bmx <- raw$P_BMX %>% select(SEQN, BMXBMI)
smq <- raw$P_SMQ %>% select(SEQN, SMQ020)
alq <- raw$P_ALQ %>% select(SEQN, ALQ111)
huq <- raw$P_HUQ %>% select(SEQN, HUQ010)
bpq <- raw$P_BPQ %>% select(SEQN, BPQ020)

merged <- demo %>%
  left_join(hiq, by = "SEQN") %>%
  left_join(diq, by = "SEQN") %>%
  left_join(bmx, by = "SEQN") %>%
  left_join(smq, by = "SEQN") %>%
  left_join(alq, by = "SEQN") %>%
  left_join(huq, by = "SEQN") %>%
  left_join(bpq, by = "SEQN")

dim(merged)
```

# 4. Recode variables

NHANES uses numeric codes with `7`/`77`/`9`/`99` reserved for *Refused* and
*Don't know*; we map those to `NA`. Treatment and outcome are binarized as
defined in the estimand.

```{r recode}
analytic_full <- merged %>%
  mutate(
    # ---- treatment: insurance coverage ----
    insured = case_when(
      HIQ011 == 1 ~ "Insured",
      HIQ011 == 2 ~ "Uninsured",
      TRUE        ~ NA_character_           # 7 = refused, 9 = don't know
    ),
    insured = factor(insured, levels = c("Uninsured", "Insured")),

    # ---- outcome: blood-sugar screening in past 3 years ----
    screened = case_when(
      DIQ180 == 1 ~ "Screened",
      DIQ180 == 2 ~ "Not screened",
      TRUE        ~ NA_character_           # 7/9 and missing
    ),
    screened = factor(screened, levels = c("Not screened", "Screened")),

    # ---- covariates ----
    age = RIDAGEYR,
    sex = factor(RIAGENDR, levels = c(1, 2), labels = c("Male", "Female")),
    race_eth = factor(RIDRETH3,
      levels = c(1, 2, 3, 4, 6, 7),
      labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White",
                 "Non-Hispanic Black", "Non-Hispanic Asian", "Other/Multi")),
    education = factor(case_when(
      DMDEDUC2 %in% 1:2 ~ "< High school",
      DMDEDUC2 == 3     ~ "High school / GED",
      DMDEDUC2 == 4     ~ "Some college",
      DMDEDUC2 == 5     ~ "College graduate",
      TRUE              ~ NA_character_),
      levels = c("< High school", "High school / GED",
                 "Some college", "College graduate")),
    marital = factor(case_when(
      DMDMARTZ == 1 ~ "Married / partnered",
      DMDMARTZ == 2 ~ "Widowed / divorced / separated",
      DMDMARTZ == 3 ~ "Never married",
      TRUE          ~ NA_character_),
      levels = c("Married / partnered",
                 "Widowed / divorced / separated", "Never married")),
    us_born = factor(case_when(
      DMDBORN4 == 1 ~ "Born in U.S.",
      DMDBORN4 == 2 ~ "Born outside U.S.",
      TRUE          ~ NA_character_),
      levels = c("Born in U.S.", "Born outside U.S.")),
    interview_lang = factor(case_when(
      SIALANG == 1 ~ "English",
      SIALANG == 2 ~ "Spanish",
      TRUE         ~ NA_character_),
      levels = c("English", "Spanish")),
    pir = INDFMPIR,        # family income-to-poverty ratio (0–5, topcoded)
    bmi = BMXBMI,

    # ---- health status / risk ----
    gen_health = factor(case_when(
      HUQ010 == 1 ~ "Excellent",
      HUQ010 == 2 ~ "Very good",
      HUQ010 == 3 ~ "Good",
      HUQ010 == 4 ~ "Fair",
      HUQ010 == 5 ~ "Poor",
      TRUE        ~ NA_character_),         # 7/9 = refused / don't know
      levels = c("Excellent", "Very good", "Good", "Fair", "Poor")),
    hypertension = factor(case_when(
      BPQ020 == 1 ~ "Yes",
      BPQ020 == 2 ~ "No",
      TRUE        ~ NA_character_),
      levels = c("No", "Yes")),

    # ---- health behavior ----
    smoked_100 = factor(case_when(
      SMQ020 == 1 ~ "Yes",
      SMQ020 == 2 ~ "No",
      TRUE        ~ NA_character_),
      levels = c("No", "Yes")),
    alcohol_ever = factor(case_when(
      ALQ111 == 1 ~ "Yes",
      ALQ111 == 2 ~ "No",
      TRUE        ~ NA_character_),
      levels = c("No", "Yes"))
  )

# nice labels for the summary table
labelled::var_label(analytic_full) <- list(
  age       = "Age (years)",
  sex       = "Sex",
  race_eth  = "Race / ethnicity",
  education = "Education",
  marital   = "Marital status",
  us_born   = "Country of birth",
  interview_lang = "Interview language",
  pir       = "Income-to-poverty ratio",
  bmi       = "Body-mass index (kg/m²)",
  gen_health     = "Self-rated general health",
  hypertension   = "Ever told had high blood pressure",
  smoked_100     = "Ever smoked ≥100 cigarettes",
  alcohol_ever   = "Ever had an alcoholic drink",
  insured   = "Health-insurance coverage",
  screened  = "Blood-sugar screening (past 3 yr)"
)
```

# 5. Apply the assumption-driven filters

Only now do we restrict the sample, and *only* for reasons tied to the
identifying assumptions:

1. **Target population** → adults aged ≥ 20 (the age group `DMDEDUC2` and the
   screening question are defined for).
2. **Consistency / well-defined treatment & outcome** → must have answered both
   the insurance and the screening question (non-missing $A$ and $Y$).

Covariate missingness is **not** used to drop rows here — we keep those records
so the missingness itself can be inspected before any complete-case or
imputation decision in the modeling stage.

```{r filter}
analytic <- analytic_full %>%
  filter(age >= 20,
         !is.na(insured),
         !is.na(screened))

# CONSORT-style accounting of what each restriction removes
flow <- tibble::tibble(
  step = c("Merged (all DEMO respondents)",
           "Aged >= 20",
           "+ non-missing insurance (A)",
           "+ non-missing screening (Y)  [analytic set]"),
  n = c(
    nrow(analytic_full),
    sum(analytic_full$age >= 20, na.rm = TRUE),
    sum(analytic_full$age >= 20 & !is.na(analytic_full$insured), na.rm = TRUE),
    nrow(analytic)
  )
)
flow

# write the analytic dataframe to CSV
out_csv <- "nhanes_insurance_screening_analytic.csv"
analytic_out <- analytic %>%
  select(SEQN, insured, screened,
         age, sex, race_eth, education, marital, us_born, interview_lang,
         pir, bmi, gen_health, hypertension, smoked_100, alcohol_ever,
         SDMVPSU, SDMVSTRA, WTINTPRP, WTMECPRP)
readr::write_csv(analytic_out, out_csv)
cat("Wrote", nrow(analytic_out), "rows to", out_csv, "\n")
```

# 6. Positivity check — covariate distribution by treatment arm

For the ATE to be identified, every covariate stratum must contain both
insured and uninsured respondents (positivity).
```{r tbl-summary}
tbl <- analytic %>%
  select(insured, age, sex, race_eth, education, marital, us_born,
         interview_lang, pir, bmi, gen_health, hypertension,
         smoked_100, alcohol_ever) %>%
  tbl_summary(
    by = insured,
    missing = "ifany",
    missing_text = "Missing",
    statistic = list(all_continuous()  ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1
  ) %>%
  add_overall() %>%
  add_p() %>%                                  # association with treatment
  add_n() %>%
  modify_header(label ~ "**Covariate**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Insurance status (treatment)**") %>%
  bold_labels()

tbl
```
```{r tbl-by-outcome}
# For completeness, the same distribution stratified by the OUTCOME.
analytic %>%
  select(screened, age, sex, race_eth, education, marital, us_born,
         interview_lang, pir, bmi, gen_health, hypertension,
         smoked_100, alcohol_ever) %>%
  tbl_summary(
    by = screened, missing = "ifany",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  bold_labels()
```

# 7. Exploratory covariate analysis

```{r explore-treatment-balance}
# Standardized-mean-difference-style overlap: continuous covariates by arm
analytic %>%
  select(insured, age, pir, bmi) %>%
  pivot_longer(-insured, names_to = "covariate", values_to = "value") %>%
  ggplot(aes(value, fill = insured)) +
  geom_density(alpha = 0.45) +
  facet_wrap(~covariate, scales = "free", labeller = as_labeller(
    c(age = "Age (years)", pir = "Income-to-poverty ratio",
      bmi = "BMI (kg/m²)"))) +
  labs(title = "Overlap of continuous covariates across treatment arms",
       x = NULL, y = "Density", fill = NULL) +
  theme(legend.position = "top")
```
```{r explore-categorical-overlap, fig.height=6}
# Categorical covariate composition within each treatment arm
analytic %>%
  select(insured, sex, race_eth, education, marital, us_born,
         interview_lang, gen_health, hypertension, smoked_100, alcohol_ever) %>%
  mutate(across(-insured, as.character)) %>%
  pivot_longer(-insured, names_to = "covariate", values_to = "level") %>%
  filter(!is.na(level)) %>%
  count(covariate, insured, level) %>%
  group_by(covariate, insured) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(prop, level, fill = insured)) +
  geom_col(position = "dodge") +
  facet_wrap(~covariate, scales = "free_y", ncol = 2) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Categorical covariate distribution within each treatment arm",
       x = "Within-arm proportion", y = NULL, fill = NULL) +
  theme(legend.position = "top")
```
```{r explore-missingness}
# Covariate missingness (informs complete-case vs imputation in modeling)
analytic %>%
  summarise(across(c(age, sex, race_eth, education, marital, us_born,
                     interview_lang, pir, bmi, gen_health, hypertension,
                     smoked_100, alcohol_ever),
                   ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "covariate",
               values_to = "pct_missing") %>%
  arrange(desc(pct_missing)) %>%
  mutate(pct_missing = scales::percent(pct_missing, accuracy = 0.1))
```
```{r crude-outcome}
# Crude (unadjusted) screening rate by insurance — the naive contrast
analytic %>%
  group_by(insured) %>%
  summarise(n = n(),
            screened_rate = mean(screened == "Screened"),
            .groups = "drop")
```




# 8. Session info

```{r session}
sessionInfo()
```