---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Sensitivity analyses"
author: "Sujit Silas Armstrong, Janvi Bharucha, Amanda Millatt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("survey", "dplyr", "tidyr", "ggplot2", "EValue", "mice")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

options(survey.lonely.psu = "adjust")
theme_set(theme_minimal(base_size = 12))

# analytic dataset lives in the project root
data_path <- c("nhanes_insurance_screening_analytic.csv",
               "../nhanes_insurance_screening_analytic.csv")
data_path <- data_path[file.exists(data_path)][1]
```

# 1. Purpose

The primary analyses (outcome regression, IPW, doubly robust AIPW) identify the
ATE of insurance on blood-sugar screening under **consistency**, **conditional
exchangeability** given the DAG adjustment set, and **positivity**. These
sensitivity analyses ask how fragile that conclusion is to violations of those
assumptions and to analytic choices. Each check re-uses the **doubly robust
(AIPW)** estimator as the workhorse, since it is the primary headline method.

```{r load}
df <- read.csv(data_path, stringsAsFactors = FALSE)

df <- df %>%
  mutate(
    screened_bin = ifelse(screened == "Screened", 1, 0),
    treat        = ifelse(insured == "Insured", 1, 0),
    insured  = factor(insured, levels = c("Uninsured", "Insured")),
    sex      = factor(sex, levels = c("Male", "Female")),
    race_eth = factor(race_eth, levels = c("Non-Hispanic White", "Mexican American",
                                           "Other Hispanic", "Non-Hispanic Black",
                                           "Non-Hispanic Asian", "Other/Multi")),
    education = factor(education, levels = c("< High school", "High school / GED",
                                             "Some college", "College graduate")),
    marital  = factor(marital, levels = c("Married / partnered",
                                          "Widowed / divorced / separated",
                                          "Never married")),
    us_born  = factor(us_born, levels = c("Born in U.S.", "Born outside U.S.")),
    interview_lang = factor(interview_lang, levels = c("English", "Spanish")),
    gen_health   = factor(gen_health, levels = c("Excellent", "Very good",
                                                 "Good", "Fair", "Poor")),
    hypertension = factor(hypertension, levels = c("No", "Yes")),
    smoked_100   = factor(smoked_100,   levels = c("No", "Yes")),
    alcohol_ever = factor(alcohol_ever, levels = c("No", "Yes"))
  ) %>%
  filter(WTMECPRP > 0)

# primary adjustment set from the DAG (alcohol excluded for missingness)
primary_covs <- c("age", "sex", "race_eth", "education", "pir", "marital",
                  "us_born", "interview_lang", "bmi", "gen_health",
                  "hypertension", "smoked_100")
```

```{r aipw-helper}
# Reusable AIPW estimator. Returns the risk difference and risk ratio with
# design-based (delta-method) CIs. `ps_bounds` either truncates the propensity
# score (drop_outside = FALSE) or trims units outside the bounds
# (drop_outside = TRUE). Mirrors modeling/doubly_robust.Rmd.
aipw_estimate <- function(data, covs, ps_bounds = c(0, 1), drop_outside = FALSE) {
  d <- tidyr::drop_na(data, dplyr::all_of(c("treat", "screened_bin", covs)))
  d$sw <- d$WTMECPRP / mean(d$WTMECPRP)        # rescale weights for stable glm

  psf <- glm(reformulate(covs, "treat"), d, family = quasibinomial(), weights = sw)
  d$ps <- predict(psf, type = "response")

  if (drop_outside) {
    d <- d[d$ps >= ps_bounds[1] & d$ps <= ps_bounds[2], ]
  } else {
    d$ps <- pmin(pmax(d$ps, ps_bounds[1]), ps_bounds[2])
  }

  o1 <- glm(reformulate(covs, "screened_bin"), dplyr::filter(d, treat == 1),
            family = quasibinomial(), weights = sw)
  o0 <- glm(reformulate(covs, "screened_bin"), dplyr::filter(d, treat == 0),
            family = quasibinomial(), weights = sw)
  d$m1 <- predict(o1, d, type = "response")
  d$m0 <- predict(o0, d, type = "response")

  d$mu1 <- d$m1 + d$treat       / d$ps       * (d$screened_bin - d$m1)
  d$mu0 <- d$m0 + (1 - d$treat) / (1 - d$ps) * (d$screened_bin - d$m0)

  des <- svydesign(id = ~SDMVPSU, strata = ~SDMVSTRA, weights = ~WTMECPRP,
                   data = d, nest = TRUE)
  po <- svymean(~mu1 + mu0, des)
  rd <- svycontrast(po, quote(mu1 - mu0))
  rr <- svycontrast(po, quote(mu1 / mu0))

  data.frame(
    n     = nrow(d),
    rd    = as.numeric(coef(rd)),
    rd_se = as.numeric(SE(rd)),
    rd_lo = confint(rd)[1], rd_hi = confint(rd)[2],
    rr    = as.numeric(coef(rr)),
    rr_lo = confint(rr)[1], rr_hi = confint(rr)[2]
  )
}

# primary estimate for reference throughout
primary <- aipw_estimate(df, primary_covs)
round(primary, 4)
```

# 2. Unmeasured confounding (E-value)

How strong would an unmeasured confounder have to be — associated with *both*
insurance and screening on the risk-ratio scale — to fully explain away the
observed effect (and to push the confidence bound to the null)?

```{r evalue}
ev <- EValue::evalues.RR(est = primary$rr, lo = primary$rr_lo, hi = primary$rr_hi)
ev
```

```{r evalue-interpret, echo=FALSE, results='asis'}
e_point <- ev["E-values", "point"]
e_ci    <- ev["E-values", "lower"]
cat(sprintf(
  "An unmeasured confounder would need a risk ratio of **%.2f** with both the exposure and the outcome (beyond the measured covariates) to explain away the point estimate, and **%.2f** to move the lower confidence bound to the null.\n",
  e_point, e_ci))
```

# 3. Positivity: propensity trimming and truncation

The propensity overlap is thin at the top (adults 65+ are ~98% insured via
Medicare). We re-estimate the ATE after (a) **truncating** the propensity score
to progressively tighter bounds and (b) **trimming** the sample to the region of
common support, and check whether the conclusion is stable.

```{r trimming}
positivity <- dplyr::bind_rows(
  cbind(spec = "No trimming",            aipw_estimate(df, primary_covs)),
  cbind(spec = "Truncate PS [0.05,0.95]", aipw_estimate(df, primary_covs, c(0.05, 0.95))),
  cbind(spec = "Truncate PS [0.10,0.90]", aipw_estimate(df, primary_covs, c(0.10, 0.90))),
  cbind(spec = "Trim to PS [0.05,0.95]",  aipw_estimate(df, primary_covs, c(0.05, 0.95),
                                                        drop_outside = TRUE))
)
positivity %>%
  transmute(spec, n,
            risk_difference = round(rd, 4),
            CI = sprintf("(%.3f, %.3f)", rd_lo, rd_hi))
```

```{r trimming-plot, echo=FALSE, fig.height=3.2}
positivity %>%
  mutate(spec = factor(spec, levels = rev(spec))) %>%
  ggplot(aes(rd, spec)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(aes(xmin = rd_lo, xmax = rd_hi)) +
  labs(title = "ATE is stable across positivity restrictions",
       x = "AIPW risk difference (95% CI)", y = NULL)
```

# 4. Alternative adjustment sets

Conditional exchangeability assumes the right covariates are adjusted for. We
compare the primary set against (a) adding `alcohol_ever` back (at the cost of
~14% more missingness), (b) demographics/SES only, and (c) a minimal set.

```{r adjustment-sets}
sets <- list(
  "Primary (12 covariates)" = primary_covs,
  "+ alcohol_ever"          = c(primary_covs, "alcohol_ever"),
  "Demographic / SES only"  = c("age", "sex", "race_eth", "education", "pir",
                                "marital", "us_born", "interview_lang"),
  "Minimal (age, sex, race, educ, income)" =
                              c("age", "sex", "race_eth", "education", "pir")
)

adjustment <- dplyr::bind_rows(lapply(names(sets), function(nm) {
  cbind(adjustment_set = nm, aipw_estimate(df, sets[[nm]]))
}))
adjustment %>%
  transmute(adjustment_set, n,
            risk_difference = round(rd, 4),
            CI = sprintf("(%.3f, %.3f)", rd_lo, rd_hi))
```

# 5. Missing-data handling: complete case vs. multiple imputation

The primary analyses use complete cases (`pir` and `alcohol_ever` drive most of
the loss). Here we multiply-impute the covariates, re-estimate the AIPW risk
difference within each completed dataset, and pool with Rubin's rules — checking
whether complete-case analysis materially biases the estimate.

```{r mice, results='hide'}
set.seed(235)

# impute the covariates + treatment/outcome; keep design vars as-is
impute_vars <- c("treat", "screened_bin", primary_covs)
design_vars <- c("SDMVPSU", "SDMVSTRA", "WTMECPRP")

mids <- mice(df[, c(impute_vars, design_vars)],
             m = 5, maxit = 5, printFlag = FALSE)

# AIPW risk difference + SE in each completed dataset
mi_fits <- lapply(1:mids$m, function(i) {
  d_i <- complete(mids, i)
  aipw_estimate(d_i, primary_covs)[, c("rd", "rd_se")]
})
mi_mat <- do.call(rbind, mi_fits)
```

```{r mice-pool}
# Rubin's rules
m       <- nrow(mi_mat)
qbar    <- mean(mi_mat$rd)                       # pooled point estimate
ubar    <- mean(mi_mat$rd_se^2)                  # within-imputation variance
b       <- var(mi_mat$rd)                        # between-imputation variance
T_var   <- ubar + (1 + 1 / m) * b               # total variance
se_pool <- sqrt(T_var)

data.frame(
  method = c("Complete case (primary)", "Multiple imputation (m=5, pooled)"),
  risk_difference = round(c(primary$rd, qbar), 4),
  CI = c(sprintf("(%.3f, %.3f)", primary$rd_lo, primary$rd_hi),
         sprintf("(%.3f, %.3f)", qbar - 1.96 * se_pool, qbar + 1.96 * se_pool))
)
```

# 6. Summary

```{r summary, echo=FALSE}
cat("Primary AIPW risk difference: ",
    sprintf("%.3f (95%% CI %.3f, %.3f)\n", primary$rd, primary$rd_lo, primary$rd_hi))
cat("\nThe effect direction and rough magnitude should persist across all four\n",
    "checks above; large movement in any one flags the assumption the headline\n",
    "estimate is most fragile to.\n")
```

# 7. Session info

```{r session}
sessionInfo()
```
