---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Sensitivity analyses"
author: "Sujit Silas Armstrong, Janvi Bharucha, Amanda Millatt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("survey", "dplyr", "tidyr", "ggplot2")  # add: EValue, mice as needed
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

options(survey.lonely.psu = "adjust")

# analytic dataset lives in the project root
data_path <- c("nhanes_insurance_screening_analytic.csv",
               "../nhanes_insurance_screening_analytic.csv")
data_path <- data_path[file.exists(data_path)][1]
```

> **STATUS: PLACEHOLDER.** Scaffolding for the robustness checks that probe the
> identifying assumptions behind the primary estimates (outcome regression, IPW,
> doubly robust). Analysis chunks are `eval=FALSE` so the file knits; fill in and
> enable as each check is built.

# 1. Purpose

The primary analyses identify the ATE under **consistency**, **conditional
exchangeability** given the DAG adjustment set, and **positivity**. These
sensitivity analyses ask how fragile the conclusion is to violations of those
assumptions and to analytic choices.

# 2. Unmeasured confounding (E-value)  *(TODO)*

How strong would an unmeasured confounder have to be — in association with both
insurance and screening — to explain away the observed effect?

```{r evalue, eval=FALSE}
# library(EValue)
# point estimate + CI bound from the primary model (risk ratio scale)
# evalues.RR(est = <RR>, lo = <CI_low>, hi = <CI_high>)
```

# 3. Positivity / extreme weights  *(TODO)*

Re-estimate after (a) trimming the propensity score to common support and
(b) truncating IPT weights at the 1st/99th percentiles.

```{r trimming, eval=FALSE}
# refit the IPW / AIPW pipeline restricting to 0.01 <= ps <= 0.99
# and compare the ATE to the untrimmed estimate.
```

# 4. Alternative adjustment sets  *(TODO)*

Compare the primary adjustment set against (a) adding `alcohol_ever` back in
(at the cost of ~14% more missingness) and (b) a minimal sufficient set implied
by the DAG.

```{r adjustment-sets, eval=FALSE}
# loop over candidate covariate sets, re-estimate, tabulate the insurance effect.
```

# 5. Missing-data handling  *(TODO)*

Compare complete-case analysis (primary) against multiple imputation for the
covariates with the most missingness (`pir`, `alcohol_ever`).

```{r missing-data, eval=FALSE}
# library(mice)
# impute L, re-estimate within each completed dataset, pool with Rubin's rules.
```

# 6. Session info

```{r session}
sessionInfo()
```
