---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Doubly robust estimation (AIPW)"
author: "Sujit Silas Armstrong"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("survey", "dplyr", "tidyr", "ggplot2")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

options(survey.lonely.psu = "adjust")
theme_set(theme_minimal(base_size = 12))

# analytic dataset lives in the project root; works whether this file is
# knit from modeling/ or run interactively from the project root
data_path <- c("nhanes_insurance_screening_analytic.csv",
               "../nhanes_insurance_screening_analytic.csv")
data_path <- data_path[file.exists(data_path)][1]
```


# 1. Overview

**Estimand.** Average treatment effect (ATE) of health-insurance coverage
($A$: insured vs. uninsured) on blood-sugar screening ($Y$), on the **risk-
difference** scale,
$$\tau = \mathbb{E}\!\left[Y^{1}\right] - \mathbb{E}\!\left[Y^{0}\right],$$
identified under the assumptions in `../nhanes_data_prep.Rmd` and the adjustment
set in `../nhanes_dag.Rmd`.

**Method — Augmented IPW (AIPW).** The doubly robust estimator combines a
**propensity-score** model $e(L)=\Pr(A=1\mid L)$ with an **outcome** model
$m_a(L)=\mathbb{E}[Y\mid A=a, L]$ through the per-subject pseudo-outcomes

$\hat\tau = \overline{\hat\mu_1} - \overline{\hat\mu_0}$ is **consistent if
*either*** the propensity model **or** the outcome model is correctly specified
(double robustness). We carry the NHANES survey weights through every step so
the estimate targets the U.S. adult population, and we obtain design-based
confidence intervals from the survey package.

# 2. Load, recode, complete-case

```{r load-recode}
df <- read.csv(data_path, stringsAsFactors = FALSE)

df <- df %>%
  mutate(
    screened_bin = ifelse(screened == "Screened", 1, 0),
    treat        = ifelse(insured == "Insured", 1, 0),   # 1 = insured (treated)

    insured  = factor(insured, levels = c("Uninsured", "Insured")),
    sex      = factor(sex, levels = c("Male", "Female")),
    race_eth = factor(race_eth, levels = c("Non-Hispanic White", "Mexican American",
                                           "Other Hispanic", "Non-Hispanic Black",
                                           "Non-Hispanic Asian", "Other/Multi")),
    education = factor(education, levels = c("< High school", "High school / GED",
                                             "Some college", "College graduate")),
    marital  = factor(marital, levels = c("Married / partnered",
                                          "Widowed / divorced / separated",
                                          "Never married")),
    us_born  = factor(us_born, levels = c("Born in U.S.", "Born outside U.S.")),
    interview_lang = factor(interview_lang, levels = c("English", "Spanish")),
    gen_health   = factor(gen_health, levels = c("Excellent", "Very good",
                                                 "Good", "Fair", "Poor")),
    hypertension = factor(hypertension, levels = c("No", "Yes")),
    smoked_100   = factor(smoked_100,   levels = c("No", "Yes")),
    alcohol_ever = factor(alcohol_ever, levels = c("No", "Yes"))
  ) %>%
  filter(WTMECPRP > 0)            # keep the MEC-examined sample (BMI measured there)

# adjustment set L (from the DAG). alcohol is dropped here to limit the
# complete-case loss it drives (~14% missing); revisit in sensitivity analysis.
covariates <- c("age", "sex", "race_eth", "education", "pir", "marital",
                "us_born", "interview_lang", "bmi", "gen_health",
                "hypertension", "smoked_100")

model_df <- df %>%
  tidyr::drop_na(all_of(c("treat", "screened_bin", covariates)))

cat("MEC-examined n:      ", nrow(df), "\n")
cat("complete-case n (L): ", nrow(model_df), "\n")
cat("  insured:   ", sum(model_df$treat == 1), "\n")
cat("  uninsured: ", sum(model_df$treat == 0), "\n")
```

# 3. Nuisance models

Both nuisance models are fit with the NHANES exam weights so they estimate
population quantities. Two implementation details matter here:

- We use `quasibinomial()` to avoid spurious non-integer-weight warnings;
  coefficients are identical to `binomial()`.

```{r nuisance}
# rescale survey weights to mean 1
model_df$sw <- model_df$WTMECPRP / mean(model_df$WTMECPRP)

# ---- propensity score: e(L) = P(insured = 1 | L) ----
ps_formula <- reformulate(covariates, response = "treat")
ps_fit <- glm(ps_formula, data = model_df,
              family = quasibinomial(), weights = sw)
model_df$ps <- predict(ps_fit, type = "response")

# ---- outcome models: m_a(L) = E(Y | A = a, L), fit separately by arm ----
y_formula <- reformulate(covariates, response = "screened_bin")
out1 <- glm(y_formula, data = filter(model_df, treat == 1),
            family = quasibinomial(), weights = sw)
out0 <- glm(y_formula, data = filter(model_df, treat == 0),
            family = quasibinomial(), weights = sw)

# counterfactual predictions for everyone under each arm
model_df$m1 <- predict(out1, newdata = model_df, type = "response")
model_df$m0 <- predict(out0, newdata = model_df, type = "response")
```

## 3.1 Positivity / overlap diagnostics

```{r overlap, fig.height=4}
cat("propensity range, uninsured:",
    round(range(model_df$ps[model_df$treat == 0]), 4), "\n")
cat("propensity range, insured:  ",
    round(range(model_df$ps[model_df$treat == 1]), 4), "\n")

ggplot(model_df, aes(ps, fill = insured)) +
  geom_density(alpha = 0.45) +
  labs(title = "Propensity-score overlap by insurance status",
       subtitle = "Positivity holds where both arms have density across the PS range",
       x = "Estimated P(insured | L)", y = "Density", fill = NULL) +
  theme(legend.position = "top")

# IPT weights implied by the PS — inspect the tails for instability
model_df$ipt_w <- ifelse(model_df$treat == 1, 1 / model_df$ps, 1 / (1 - model_df$ps))
cat("\nIPT weight summary (unstabilized):\n"); print(summary(model_df$ipt_w))
```

# 4. AIPW estimator

```{r aipw}
model_df <- model_df %>%
  mutate(
    mu1 = m1 + treat       / ps       * (screened_bin - m1),   # pseudo-outcome for E[Y^1]
    mu0 = m0 + (1 - treat) / (1 - ps) * (screened_bin - m0),   # pseudo-outcome for E[Y^0]
    aipw_contrast = mu1 - mu0
  )

# design-based inference: the survey-weighted mean of the pseudo-outcomes IS the
# AIPW estimate of each potential-outcome mean; svycontrast gives the RD/RR/OR
# with delta-method CIs that respect the NHANES clustering and stratification.
aipw_design <- svydesign(id = ~SDMVPSU, strata = ~SDMVSTRA,
                         weights = ~WTMECPRP, data = model_df, nest = TRUE)

po  <- svymean(~mu1 + mu0, aipw_design)            # E[Y^1], E[Y^0]
rd  <- svycontrast(po, quote(mu1 - mu0))                           # risk difference
rr  <- svycontrast(po, quote(mu1 / mu0))                           # risk ratio
or  <- svycontrast(po, quote((mu1 / (1 - mu1)) / (mu0 / (1 - mu0))))  # marginal OR

aipw_results <- rbind(
  `E[Y^1] (insured)`   = c(coef(po)["mu1"], confint(po)["mu1", ]),
  `E[Y^0] (uninsured)` = c(coef(po)["mu0"], confint(po)["mu0", ]),
  `Risk difference`    = c(coef(rd), confint(rd)),
  `Risk ratio`         = c(coef(rr), confint(rr)),
  `Marginal OR`        = c(coef(or), confint(or))
)
colnames(aipw_results) <- c("estimate", "ci_low", "ci_high")
round(aipw_results, 4)
```

The **risk difference** is the headline causal estimate: the change in the
population probability of blood-sugar screening attributable to insurance
coverage. The **marginal OR** is reported for comparison with the conditional
ORs from the outcome-regression document (`outcome_regression.Rmd`); marginal
and conditional ORs are not expected to be equal (non-collapsibility).

# 5. Double robustness — the two halves vs. AIPW

The AIPW estimate should sit close to its two ingredients (plug-in g-formula and
IPW) and is the one that stays valid if *either* nuisance model is wrong.

```{r dr-comparison}
w <- model_df$WTMECPRP

# (a) plug-in g-formula (outcome model only)
rd_g <- weighted.mean(model_df$m1 - model_df$m0, w)

# (b) IPW / Hajek (propensity model only)
ey1_ipw <- with(model_df, sum(w * treat * screened_bin / ps) /
                          sum(w * treat / ps))
ey0_ipw <- with(model_df, sum(w * (1 - treat) * screened_bin / (1 - ps)) /
                          sum(w * (1 - treat) / (1 - ps)))
rd_ipw <- ey1_ipw - ey0_ipw

# (c) AIPW (doubly robust)
rd_aipw <- weighted.mean(model_df$aipw_contrast, w)

data.frame(
  estimator = c("g-formula (outcome only)", "IPW (propensity only)",
                "AIPW (doubly robust)"),
  risk_difference = round(c(rd_g, rd_ipw, rd_aipw), 4)
)
```

# 6. Alternative inference (treats nuisances as fixed, iid)

A simpler influence-function SE that ignores the survey design, shown only as a
cross-check on the design-based CI above. The design-based result in Section 4
is the one we report.

```{r if-se}
n   <- nrow(model_df)
psi <- mean(model_df$aipw_contrast)                 # unweighted sample ATE
se  <- sd(model_df$aipw_contrast) / sqrt(n)
cat(sprintf("iid AIPW risk difference: %.4f  (95%% CI %.4f to %.4f)\n",
            psi, psi - 1.96 * se, psi + 1.96 * se))
```

# 7. Session info

```{r session}
sessionInfo()
```