---
title: "NHANES 2017-2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Inverse Probability Weighting (IPW)"
author: "Amanda Millatt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup-env, fig.width=10, fig.height=8}

#Loading and bringing in the required packages

pkgs <- c("cobalt", "survey", "dplyr", "ggplot2", "boot")

to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

#Pulling in the final dataset 

df <- read.csv("nhanes_insurance_screening_analytic.csv")

theme_set(theme_minimal(base_size = 12))
```

# 1.Estimating propensity score through a logistic regression model 

The propensity score, which is the probability of treatment, is calculated by first regressing the *insurance status* on the covariates identified from DAG (included the covariates below for reference). All covariates identified were confounders in the causal pathway from *insurance status* to *blood sugar screening*. 

- *Demographic / socioeconomic:* age, sex, race/ethnicity, education, family
  income-to-poverty ratio, marital status, country of birth, interview
  language.
- *Health status / risk:* BMI, self-rated general health, prior hypertension
  diagnosis. 
- *Health behavior:* ever-drinking alcohol, ever-smoking.

```{r PS}
ps_model <- glm(
  I(insured == "Insured") ~ age + sex + race_eth + education + 
                           pir + marital + us_born + interview_lang + 
                           bmi + gen_health + hypertension + smoked_100 + alcohol_ever,
  data = df,
  family = binomial(link = "logit"),
  na.action = na.exclude
)
df$propensity_score <- predict(ps_model, type = "response")

summary(df$propensity_score)
```
# 2. Checking propensity score overlap by levels of Insurance Status 

To verify that the propensity score found enough matches between the insured and uninsured group, a density plot of the propensity scores were created to visualize the overlapping areas. 

*Analysis of the curve:* The plot below shows a huge spike close to 1 for the insured group indicating that a large group of units are "near-certain" to be insured (propensity Score > 0.9). These individuals are extremely likely to have health insurance regardless of their baseline characteristics.While, the uninsured group showed a decline in propensity score density near 0.95 -1.This means there are very few uninsured units who are similar to the units in that insured spike. Due to the extreme propensity scores, the estimated ATE will be highly dependent on the regression model’s extrapolation causing a slightly biased ATE estimate. 

*Conclusion*: Covariate balance will be evaluated after applying stabilized IPW weights using a love plot, which will indicate if further methods such as as Overlap Weighting (OW) or Covariate Balancing Propensity Score (CBPS). 


```{r cars}
ggplot(df, aes(x = propensity_score, fill = insured)) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Propensity Scores",
       x = "Propensity Score (Probability of being Insured)",
       y = "Density") +
  theme_minimal()
```

# 3. Calculating Stabilized IPW Weights

To estimate ATE  while adjusting for covariates, stabilized inverse probability weights were employed. Unstabilized IPW weights can often be highly variable, especially when propensity scores are extreme, which may lead to biased estimates with large variance. Stabilized weights mitigate this by introducing the marginal probability of treatment ($P(T=1)$) into the numerator. In turn, the stabilized weights will reduce variance and ensure the weighted sample size is consistent with the orginal sample. 

```{r IPW}
df$ps <- predict(ps_model, type = "response")

df <- df%>%
  mutate(
    ipw = ifelse(insured == "Insured", 
                 mean(insured == "Insured") / ps, 
                 (1 - mean(insured == "Insured")) / (1 - ps))
  )

ipw_summary <- df%>%
  group_by(insured) %>%
  summarise(
    "N (Sample Size)" = n(),
    "Mean Stabilized Weight" = mean(ipw, na.rm = TRUE),
    "Min Weight" = min(ipw, na.rm = TRUE),
    "Max Weight" = max(ipw, na.rm = TRUE)
  )

ipw_summary 
```

# 4. Intergrating sampling weights from NHANES and fitting the Outcome Model using a logistic regression with a binomial distribution 

To run the regression model, the NHANES data had to be filtered for complete rows and then the final weights were calculated by integrating the sampling design weights with the calculated stabilized IPW weights. The  *Screened* and *Insured* variables were converted into a 0/1 numeric format compatible with logistic regression. Next, the svydesign function constructs a complex survey design object using NHANES specific stratification and cluster variables to ensure statistically valid inferences. Finally, the binomial logistic regression model was fitted to estimate the effect of insurance status on diabetes screening. The dispersion parameter was checked to ensure there was no over dispersion from the binomial distribution. The dispersion parameter was  1.000181, which indicates that the use of the binomial distribution was appropriate. 

```{r regression}
df_clean <- df[!is.na(df$ipw), ]

df_clean $final_weight <- df_clean $WTMECPRP * df_clean$ipw

#Changing the variables from string to numeric to run the regression model 
df_clean$screened_recoded <- ifelse(df_clean$screened == "Screened", 1, 0)

df_clean$insured_recoded <- ifelse(df_clean$insured == "Insured", 1, 0)

nhanes_ipw_design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~final_weight,
  data    = df_clean ,
  nest    = TRUE
)


#fitting the regression model with modified weights to account for the complexity of the sampling design weights calculated by NHANES
m_ipw <- svyglm(
  screened_recoded ~ insured_recoded,
  design = nhanes_ipw_design,
  family = binomial(link = "logit")
)

summary(m_ipw)

#Check for over dispersion
m_summary <- summary(m_ipw)
dispersion_param <- m_summary$dispersion

cat("The estimated dispersion parameter is:", dispersion_param, "\n")


```
# 5. Creating the love plot to examine covariate balance before and after adjustment 

A love plot was created to examine the covariate blanace by levels of insurance coverage with the ideal absolute standardized mean differences (SMD) threshold set at 0.1. The love plot demonstrates the effectiveness of inverse probability weighting in balancing the covariate with the absolute SMD for most variables reduced to below the 0.1 threshold after adjustment. This indicates  improved comparability between the insured and uninsured groups.

Please note: Although the SMD improved for all variables, *pir*, *education_College graduate*,*education_< High school*  and *race_eth_Non-Hispanic White* were all above the 0.1 threshold. 
```{r loveplot}

love.plot(
  insured_recoded ~ age + sex + race_eth + education + pir + bmi + hypertension + smoked_100, 
  data = df_clean,
  weights = df_clean$final_weight, 
  binary = "std",                  
  threshold = 0.1,                 
  abs = TRUE,                      
  var.order = "unadjusted"         
)

```


# 6. Calculating the ATE 

The Average Treatment Effect (ATE) was calculated by predicting the probability of diabetes screening for every unit under both insured and uninsured scenarios (potential outcomes) based on the previously fitted binomial logistic regression model.Then, the absolute difference of screening probability for each unit in the sample was estimated by subtracting these units predictions. The original NHANES survey weights **WTMECPRP** to these differences to ensure the estimate was nationally representative of the U.S. population. Finally, the weighted ATE was computed.

```{r ATE}
pred_ins   <- predict(m_ipw, newdata = mutate(df_clean, insured_recoded = 1),   type = "response")
pred_unins <- predict(m_ipw, newdata = mutate(df_clean, insured_recoded = 0), type = "response")

# Weighted ATE
ate_diff <- (pred_ins - pred_unins) * df_clean$WTMECPRP
ate_ipw  <- sum(ate_diff, na.rm = TRUE) / sum(df_clean$WTMECPRP, na.rm = TRUE)

cat("IPW-Adjusted ATE:", round(ate_ipw, 4), "\n")
```
# 7. Calculating the 95% confidence interval from bootstrapping with replacement


```{r CI}
#Creating a loop function to calculate IPW for each iteration
boot_ate_fn <- function(data, indices) {
  d_boot <- data[indices, ]
  
  d_boot$insured_recoded  <- ifelse(d_boot$insured == "Insured", 1, 0)
  d_boot$screened_recoded <- ifelse(d_boot$screened == "Screened", 1, 0)
  
#estimate propensity score model
  
  ps_m_boot <- glm(
    insured_recoded ~ age + sex + race_eth + education + 
                      pir + marital + us_born + interview_lang + 
                      bmi + gen_health + hypertension + smoked_100 + alcohol_ever,
    data = d_boot,
    family = binomial(link = "logit"),
    na.action = na.exclude
  )
  d_boot$ps <- predict(ps_m_boot, type = "response")
  
#calculate stabilized IPW weights 
  p_insured_boot <- mean(d_boot$insured_recoded == 1, na.rm = TRUE)
  d_boot <- d_boot %>%
    mutate(
      ipw = ifelse(insured_recoded == 1, 
                   p_insured_boot / ps, 
                   (1 - p_insured_boot) / (1 - ps))
    )
  
#Filter for out the blank values and merge with NHANES sampling design weights
  d_boot_clean <- d_boot[!is.na(d_boot$ipw), ]
  d_boot_clean$final_weight <- d_boot_clean$WTMECPRP * d_boot_clean$ipw
  
#Scale the weights to protect standard glm
  d_boot_clean$final_weight_scaled <- d_boot_clean$final_weight / mean(d_boot_clean$final_weight, na.rm = TRUE)
  
#Fit outcome model using scaled weights
  m_boot <- glm(
    screened_recoded ~ insured_recoded,
    data = d_boot_clean,
    weights = final_weight_scaled,
    family = binomial(link = "logit")
  )
  
#Predict potential outcomes and compute ATE
  p_ins   <- predict(m_boot, newdata = mutate(d_boot_clean, insured_recoded = 1), type = "response")
  p_unins <- predict(m_boot, newdata = mutate(d_boot_clean, insured_recoded = 0), type = "response")
  
  boot_ate_diff <- (p_ins - p_unins) * d_boot_clean$WTMECPRP
  boot_ate_ipw  <- sum(boot_ate_diff, na.rm = TRUE) / sum(d_boot_clean$WTMECPRP, na.rm = TRUE)
  
  return(boot_ate_ipw)
}

#Run the bootstrap (500 iterations)
set.seed(123) 
boot_results <- boot(data = df, statistic = boot_ate_fn, R = 500)

#Extract the 95% Confidence Interval from the bootstrapped sample
boot_ci <- boot.ci(boot_results, type = "perc")
cat("Bootstrapped 95% Confidence Interval [", round(boot_ci$percent[4], 4), ",", round(boot_ci$percent[5], 4), "]\n")
```
