---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Inverse-probability-of-treatment weighting (IPTW)"
author: "Amanda Millatt"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("survey", "dplyr", "tidyr", "ggplot2")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

options(survey.lonely.psu = "adjust")

# analytic dataset lives in the project root; works whether this file is
# knit from modeling/ or run interactively from the project root
data_path <- c("nhanes_insurance_screening_analytic.csv",
               "../nhanes_insurance_screening_analytic.csv")
data_path <- data_path[file.exists(data_path)][1]
```

# 1. Overview

**Estimand.** Average treatment effect (ATE) of health-insurance coverage on
blood-sugar screening, identified under the assumptions documented in
`../nhanes_data_prep.Rmd` and the DAG in `../nhanes_dag.Rmd`.

**Method.** Inverse-probability-of-treatment weighting:

1. Model the propensity score $e(L)=\Pr(A=1\mid L)$ with logistic regression.
2. Form stabilized IPT weights $sw = \frac{\Pr(A)}{e(L)}$ for the treated and
   $\frac{\Pr(A{=}0)}{1-e(L)}$ for the untreated.
3. Combine the IPT weights with the NHANES survey weights.
4. Estimate the marginal risk difference / odds ratio in the pseudo-population.

# 2. Load and recode (shared with the team)

```{r load-recode}
df <- read.csv(data_path, stringsAsFactors = FALSE)

df <- df %>% 
  mutate(
    screened_bin = ifelse(screened == "Screened", 1, 0),
    treat        = ifelse(insured == "Insured", 1, 0),   # 1 = insured

    insured  = factor(insured, levels = c("Uninsured", "Insured")),
    sex      = factor(sex, levels = c("Male", "Female")),
    race_eth = factor(race_eth, levels = c("Non-Hispanic White", "Mexican American",
                                           "Other Hispanic", "Non-Hispanic Black",
                                           "Non-Hispanic Asian", "Other/Multi")),
    education = factor(education, levels = c("< High school", "High school / GED",
                                             "Some college", "College graduate")),
    marital  = factor(marital, levels = c("Married / partnered",
                                          "Widowed / divorced / separated",
                                          "Never married")),
    us_born  = factor(us_born, levels = c("Born in U.S.", "Born outside U.S.")),
    interview_lang = factor(interview_lang, levels = c("English", "Spanish")),
    gen_health   = factor(gen_health, levels = c("Excellent", "Very good",
                                                 "Good", "Fair", "Poor")),
    hypertension = factor(hypertension, levels = c("No", "Yes")),
    smoked_100   = factor(smoked_100,   levels = c("No", "Yes")),
    alcohol_ever = factor(alcohol_ever, levels = c("No", "Yes"))
  ) %>%
  filter(WTMECPRP > 0)

# confounder set L (adjustment set from the DAG)
covariates <- c("age", "sex", "race_eth", "education", "pir", "marital",
                "us_born", "interview_lang", "bmi", "gen_health",
                "hypertension", "smoked_100")

# complete cases on (A, Y, L) for the weighting model
model_df <- df %>%
  tidyr::drop_na(all_of(c("treat", "screened_bin", covariates)))

cat("complete-case n:", nrow(model_df), "\n")
```

# 3. Propensity-score model  *(TODO)*

```{r ps-model, eval=FALSE}
# TODO: fit the propensity score e(L) = P(insured = 1 | L)
ps_formula <- reformulate(covariates, response = "treat")

ps_fit <- glm(ps_formula, data = model_df, family = binomial())
model_df$ps <- predict(ps_fit, type = "response")

# overlap / positivity diagnostic: PS distribution by treatment arm
ggplot(model_df, aes(ps, fill = insured)) +
  geom_density(alpha = 0.45) +
  labs(title = "Propensity-score overlap by insurance status",
       x = "Estimated P(insured | L)", y = "Density", fill = NULL)
```

# 4. Stabilized IPT weights  *(TODO)*

```{r ipt-weights, eval=FALSE}
# TODO: build stabilized IPT weights and combine with the NHANES survey weight
p_treat <- mean(model_df$treat)

model_df <- model_df %>%
  mutate(
    sw = ifelse(treat == 1, p_treat / ps, (1 - p_treat) / (1 - ps)),
    # combine with NHANES exam weight for a design-consistent estimate
    final_w = sw * WTMECPRP
  )

# check weights: mean of stabilized weights should be ~1; inspect the tail
summary(model_df$sw)
# TODO: consider truncating extreme weights (e.g. 1st/99th percentile)
```

# 5. Weighted effect estimate  *(TODO)*

```{r ipw-estimate, eval=FALSE}
# TODO: estimate the marginal effect in the IPT-weighted pseudo-population.
# Use a survey design with the combined weights so SEs respect the NHANES
# clustering/stratification.
ipw_design <- svydesign(
  id = ~SDMVPSU, strata = ~SDMVSTRA, weights = ~final_w,
  data = model_df, nest = TRUE
)

# marginal risk difference
svyby(~screened_bin, ~insured, ipw_design, svymean)

# marginal model (msm)
msm <- svyglm(screened_bin ~ insured, design = ipw_design,
              family = quasibinomial())
summary(msm)
exp(cbind(OR = coef(msm), confint(msm, df = Inf)))
```

# 6. Session info

```{r session}
sessionInfo()
```