---
title: "Health Insurance and Blood Sugar Screening: Logistic Regression Analysis"
subtitle: "NHANES 2017–2020 Pre-Pandemic Data"
author: "Janvi Bharucha"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

---

# Overview

This analysis examines whether having health insurance is associated with receiving a blood sugar screening test, using NHANES 2017–2020 pre-pandemic data. We fit three survey-weighted logistic regression models with increasing levels of covariate adjustment, and estimate the Average Treatment Effect (ATE) using g-computation (standardization).

- **Exposure:** Health insurance coverage (Insured vs. Uninsured)
- **Outcome:** Blood sugar screened in the past 3 years (Yes vs. No)
- **Estimand:** Average Treatment Effect (ATE) on the probability of screening

---

# Setup

```{r libraries}
library(survey)
library(dplyr)
library(tidyr)
library(knitr)

# suppress the common kableExtra dependency warning gracefully
use_kable_extra <- requireNamespace("kableExtra", quietly = TRUE)
if (use_kable_extra) library(kableExtra)

# tells R how to handle strata with only one PSU, which comes up in NHANES
options(survey.lonely.psu = "adjust")
```

---

# 1. Load and Prepare Data

```{r load-data}
df <- read.csv("nhanes_insurance_screening_analytic.csv", stringsAsFactors = FALSE)
```

```{r recode}
# set reference levels so model output makes sense:
# uninsured = reference for insurance, non-hispanic white = reference for race, etc.

df <- df %>%
  mutate(
    # make outcome 0/1 numeric for svyglm
    screened_bin = ifelse(screened == "Screened", 1, 0),

    # exposure — uninsured is reference
    insured = factor(insured, levels = c("Uninsured", "Insured")),

    # demographics
    sex      = factor(sex, levels = c("Male", "Female")),
    race_eth = factor(race_eth,
                      levels = c("Non-Hispanic White", "Mexican American",
                                 "Other Hispanic", "Non-Hispanic Black",
                                 "Non-Hispanic Asian", "Other/Multi")),
    education = factor(education,
                       levels = c("< High school", "High school / GED",
                                  "Some college", "College graduate")),
    marital = factor(marital,
                     levels = c("Married / partnered",
                                "Widowed / divorced / separated",
                                "Never married")),
    us_born        = factor(us_born,
                            levels = c("Born in U.S.", "Born outside U.S.")),
    interview_lang = factor(interview_lang, levels = c("English", "Spanish")),

    # health status
    gen_health   = factor(gen_health,
                          levels = c("Excellent", "Very good", "Good", "Fair", "Poor")),
    hypertension = factor(hypertension, levels = c("No", "Yes")),

    # behaviors
    smoked_100   = factor(smoked_100,   levels = c("No", "Yes")),
    alcohol_ever = factor(alcohol_ever, levels = c("No", "Yes"))
  )
```

---

# 2. Sample Flow

Respondents with a zero MEC weight did not attend the physical exam. We exclude them because BMI (from the exam) is a covariate in Model 3, and we want a consistent analytic sample across all models.

```{r sample-flow}
n_full <- nrow(df)
df_analytic <- df %>% filter(WTMECPRP > 0)
n_analytic  <- nrow(df_analytic)

flow <- data.frame(
  Step = c(
    "Full dataset",
    "Excluded: zero MEC weight (interview only, no exam)",
    "Final analytic sample"
  ),
  N = c(
    n_full,
    n_full - n_analytic,
    n_analytic
  )
)

kable(flow, caption = "Sample Flow", align = c("l", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

**Final sample breakdown:**

- Insured: **`r sum(df_analytic$insured == "Insured")`**
- Uninsured: **`r sum(df_analytic$insured == "Uninsured")`**
- Screened: **`r sum(df_analytic$screened_bin == 1, na.rm = TRUE)`**
- Not screened: **`r sum(df_analytic$screened_bin == 0, na.rm = TRUE)`**

---

# 3. Covariate Missingness

```{r missingness}
missing_check <- df_analytic %>%
  summarise(across(c(age, sex, race_eth, education, marital, us_born,
                     interview_lang, pir, bmi, gen_health,
                     hypertension, smoked_100, alcohol_ever),
                   ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N Missing") %>%
  mutate(`% Missing` = round(`N Missing` / nrow(df_analytic) * 100, 1)) %>%
  arrange(desc(`N Missing`))

kable(missing_check,
      caption = "Covariate Missingness in Analytic Sample",
      align   = c("l", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

> Income-to-poverty ratio (`pir`) and alcohol use (`alcohol_ever`) have the most missing values. Complete case analysis is used throughout — `svyglm()` automatically drops observations missing any model variable.

---

# 4. Survey Design

```{r survey-design}
# incorporates NHANES complex sample design:
# id = PSU (cluster), strata = sampling stratum, weights = MEC exam weight

nhanes_design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~WTMECPRP,
  data    = df_analytic,
  nest    = TRUE
)
```

---

# 5. Crude Screening Rates

Before modeling, we look at the raw weighted screening rates by insurance status.

```{r crude-rates}
crude_rates <- svyby(~screened_bin, ~insured, nhanes_design, svymean, na.rm = TRUE)

crude_table <- data.frame(
  `Insurance Status` = c("Uninsured", "Insured"),
  `Screening Rate (%)` = round(crude_rates$screened_bin * 100, 1),
  `Std. Error (%)` = round(crude_rates$se * 100, 1)
)

kable(crude_table,
      caption = "Weighted Blood Sugar Screening Rates by Insurance Status",
      align   = c("l", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

---

# 6. Dispersion Check

Before committing to a model family, we fit a quasibinomial to check the dispersion parameter. If it is close to 1, standard binomial is appropriate.

```{r dispersion-check}
dispersion_check <- svyglm(
  screened_bin ~ insured + age + sex + race_eth +
    education + pir + marital + us_born + interview_lang,
  design = nhanes_design,
  family = quasibinomial(link = "logit")
)

disp <- round(summary(dispersion_check)$dispersion, 4)
```

**Dispersion parameter: `r disp`**

> A dispersion parameter of `r disp` is very close to 1, confirming no meaningful overdispersion. We use standard `binomial` for all three models going forward.

---

# 7. Logistic Regression Models

## Model 1: Unadjusted

Insurance only — gives us the raw association before accounting for any covariates.

```{r model1}
m1 <- svyglm(
  screened_bin ~ insured,
  design = nhanes_design,
  family = binomial(link = "logit")
)

m1_results <- as.data.frame(exp(cbind(OR = coef(m1), confint(m1, df = Inf))))
m1_results$p_value <- round(summary(m1)$coefficients[, "Pr(>|t|)"], 4)
m1_results <- round(m1_results, 3)
m1_results$Variable <- rownames(m1_results)
rownames(m1_results) <- NULL
m1_results <- m1_results[, c("Variable", "OR", "2.5 %", "97.5 %", "p_value")]
colnames(m1_results) <- c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value")

kable(m1_results,
      caption = "Model 1: Unadjusted — Odds Ratios (95% CI)",
      align   = c("l", "r", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

## Model 2: Demographically Adjusted

Adds age, sex, race/ethnicity, education, income-to-poverty ratio, marital status, country of birth, and interview language.

```{r model2}
m2 <- svyglm(
  screened_bin ~ insured + age + sex + race_eth +
    education + pir + marital + us_born + interview_lang,
  design = nhanes_design,
  family = binomial(link = "logit")
)

m2_results <- as.data.frame(exp(cbind(OR = coef(m2), confint(m2, df = Inf))))
m2_results$p_value <- round(summary(m2)$coefficients[, "Pr(>|t|)"], 4)
m2_results <- round(m2_results, 3)
m2_results$Variable <- rownames(m2_results)
rownames(m2_results) <- NULL
m2_results <- m2_results[, c("Variable", "OR", "2.5 %", "97.5 %", "p_value")]
colnames(m2_results) <- c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value")

kable(m2_results,
      caption = "Model 2: Demographically Adjusted — Odds Ratios (95% CI)",
      align   = c("l", "r", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

## Model 3: Fully Adjusted

Adds BMI, general health, hypertension, smoking history, and alcohol use on top of Model 2.

> **Note on confidence intervals:** The masked NHANES PSU structure leaves very few residual degrees of freedom with this many covariates, causing the CIs to be unreliable (0 to Inf). We use `df = Inf` as a normal approximation workaround. The point estimates and p-values remain valid.

```{r model3}
m3 <- svyglm(
  screened_bin ~ insured + age + sex + race_eth +
    education + pir + marital + us_born + interview_lang +
    bmi + gen_health + hypertension + smoked_100 + alcohol_ever,
  design = nhanes_design,
  family = binomial(link = "logit")
)

m3_results <- as.data.frame(exp(cbind(OR = coef(m3), confint(m3, df = Inf))))
m3_results$p_value <- round(summary(m3)$coefficients[, "Pr(>|t|)"], 4)
m3_results <- round(m3_results, 3)
m3_results$Variable <- rownames(m3_results)
rownames(m3_results) <- NULL
m3_results <- m3_results[, c("Variable", "OR", "2.5 %", "97.5 %", "p_value")]
colnames(m3_results) <- c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value")

kable(m3_results,
      caption = "Model 3: Fully Adjusted — Odds Ratios (95% CI)",
      align   = c("l", "r", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

---

# 8. Insurance Effect Across Models

```{r insurance-summary}
get_insurance_row <- function(model, label) {
  or_ci <- exp(cbind(OR = coef(model), confint(model, df = Inf)))["insuredInsured", ]
  pval  <- summary(model)$coefficients["insuredInsured", "Pr(>|t|)"]
  data.frame(
    Model    = label,
    OR       = round(or_ci["OR"],    3),
    CI_lower = round(or_ci["2.5 %"], 3),
    CI_upper = round(or_ci["97.5 %"],3),
    p_value  = round(pval, 4)
  )
}

insurance_summary <- rbind(
  get_insurance_row(m1, "Unadjusted"),
  get_insurance_row(m2, "Demographically Adjusted"),
  get_insurance_row(m3, "Fully Adjusted")
)
rownames(insurance_summary) <- NULL
colnames(insurance_summary) <- c("Model", "OR", "95% CI Lower", "95% CI Upper", "p-value")

kable(insurance_summary,
      caption = "Insurance Effect (Insured vs. Uninsured) Across All Three Models",
      align   = c("l", "r", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

---

# 9. Average Treatment Effect (ATE)

The regression odds ratios describe the association on the log-odds scale, conditional on all covariates. The ATE answers a more interpretable causal question: **if we could give everyone insurance versus give no one insurance, how much would the probability of screening change on average?**

We estimate the ATE using g-computation (standardization):

1. Predict each person's probability of screening if they were **insured**
2. Predict each person's probability of screening if they were **uninsured**
3. Take the survey-weighted average of each, then subtract

```{r ate-function}
compute_ate <- function(model, weight_var = "WTMECPRP") {

  model_data <- model$data

  # counterfactual datasets: everyone insured vs. everyone uninsured
  data_insured           <- model_data
  data_uninsured         <- model_data
  data_insured$insured   <- factor("Insured",   levels = c("Uninsured", "Insured"))
  data_uninsured$insured <- factor("Uninsured", levels = c("Uninsured", "Insured"))

  # predicted probabilities under each scenario
  p_insured   <- predict(model, newdata = data_insured,   type = "response")
  p_uninsured <- predict(model, newdata = data_uninsured, type = "response")

  # align weights to the complete cases the model actually used
  used_rows <- as.integer(rownames(model$data))
  weights   <- df_analytic[[weight_var]][used_rows]

  # survey-weighted means
  mean_p_insured   <- sum(weights * p_insured,   na.rm = TRUE) / sum(weights, na.rm = TRUE)
  mean_p_uninsured <- sum(weights * p_uninsured, na.rm = TRUE) / sum(weights, na.rm = TRUE)

  list(
    p_if_all_insured   = round(mean_p_insured,   4),
    p_if_all_uninsured = round(mean_p_uninsured, 4),
    ATE                = round(mean_p_insured - mean_p_uninsured, 4)
  )
}
```

```{r ate-results}
ate1 <- compute_ate(m1)
ate2 <- compute_ate(m2)
ate3 <- compute_ate(m3)

ate_table <- data.frame(
  Model = c("Unadjusted", "Demographically Adjusted", "Fully Adjusted"),
  `P(Screened | All Insured)`   = c(ate1$p_if_all_insured,
                                     ate2$p_if_all_insured,
                                     ate3$p_if_all_insured),
  `P(Screened | All Uninsured)` = c(ate1$p_if_all_uninsured,
                                     ate2$p_if_all_uninsured,
                                     ate3$p_if_all_uninsured),
  ATE = c(ate1$ATE, ate2$ATE, ate3$ATE)
)
colnames(ate_table) <- c("Model",
                          "P(Screened | All Insured)",
                          "P(Screened | All Uninsured)",
                          "ATE")

kable(ate_table,
      caption = "Average Treatment Effect (ATE) Across Models — G-Computation",
      align   = c("l", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

> **Interpretation (Model 2):** If everyone in the population had health insurance, the probability of receiving a blood sugar screening would be approximately **`r round((ate2$ATE)*100, 1)` percentage points higher** than if no one had insurance, after adjusting for sociodemographic differences.

---

# 10. Wald Tests for Insurance Term

Formal test of whether the insurance term contributes significantly to each model beyond chance.

```{r wald-tests}
w1 <- regTermTest(m1, ~insured)
w2 <- regTermTest(m2, ~insured)
w3 <- regTermTest(m3, ~insured)

wald_table <- data.frame(
  Model   = c("Unadjusted", "Demographically Adjusted", "Fully Adjusted"),
  F_stat  = round(c(w1$Ftest["F"], w2$Ftest["F"], w3$Ftest["F"]), 3),
  df1     = c(w1$Ftest["df1"], w2$Ftest["df1"], w3$Ftest["df1"]),
  df2     = c(w1$Ftest["df2"], w2$Ftest["df2"], w3$Ftest["df2"]),
  p_value = round(c(w1$p, w2$p, w3$p), 6)
)
colnames(wald_table) <- c("Model", "F Statistic", "df1", "df2", "p-value")

kable(wald_table,
      caption = "Wald Tests for Insurance Term",
      align   = c("l", "r", "r", "r", "r")) %>%
  { if (use_kable_extra) kable_styling(., bootstrap_options = c("striped", "hover"),
                                        full_width = FALSE) else . }
```

---

# 11. Save Results

```{r save-results}
write.csv(ate_table,
          "ate_results.csv", row.names = FALSE)

or_summary <- data.frame(
  Model    = c("Unadjusted", "Demographically Adjusted", "Fully Adjusted"),
  OR       = c(exp(coef(m1)["insuredInsured"]),
               exp(coef(m2)["insuredInsured"]),
               exp(coef(m3)["insuredInsured"])),
  p_value  = c(summary(m1)$coefficients["insuredInsured", "Pr(>|t|)"],
               summary(m2)$coefficients["insuredInsured", "Pr(>|t|)"],
               summary(m3)$coefficients["insuredInsured", "Pr(>|t|)"])
)
write.csv(or_summary,
          "insurance_effect_by_model.csv", row.names = FALSE)

cat("results saved to ate_results.csv and insurance_effect_by_model.csv\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
