---
title: "NHANES 2017–2020: Health Insurance and Blood-Sugar Screening"
subtitle: "Outcome regression (survey-weighted logistic models)"
author: "Janvi Bharucha"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)

# ---- packages ----
pkgs <- c("survey", "dplyr", "tidyr")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)
invisible(lapply(pkgs, library, character.only = TRUE))

# tells R how to handle strata that only have one PSU (common in NHANES)
options(survey.lonely.psu = "adjust")

# analytic dataset lives in the project root; works whether this file is
# knit from modeling/ or run interactively from the project root
data_path <- c("nhanes_insurance_screening_analytic.csv",
               "../nhanes_insurance_screening_analytic.csv")
data_path <- data_path[file.exists(data_path)][1]
```

# 1. Overview

**Question.** Does having health insurance affect whether someone gets a
blood-sugar screening?

This document estimates the association using **survey-weighted logistic
regression** (the outcome-regression / G-computation building block of the
project). It consumes the shared analytic dataset built in
`../nhanes_data_prep.Rmd` and reports the insurance odds ratio under three
levels of adjustment.

# 2. Load and recode

We set reference levels so the model output is interpretable: **uninsured** is
the reference for insurance, **non-Hispanic White** for race/ethnicity, etc.

```{r load-recode}
df <- read.csv(data_path, stringsAsFactors = FALSE)

df <- df %>%
  mutate(
    # make outcome a 0/1 numeric so svyglm is happy
    screened_bin = ifelse(screened == "Screened", 1, 0),

    # exposure - uninsured is reference
    insured = factor(insured, levels = c("Uninsured", "Insured")),

    # demographics
    sex = factor(sex, levels = c("Male", "Female")),

    race_eth = factor(race_eth, levels = c("Non-Hispanic White", "Mexican American",
                                           "Other Hispanic", "Non-Hispanic Black",
                                           "Non-Hispanic Asian", "Other/Multi")),

    education = factor(education, levels = c("< High school", "High school / GED",
                                             "Some college", "College graduate")),

    marital = factor(marital, levels = c("Married / partnered",
                                         "Widowed / divorced / separated",
                                         "Never married")),

    us_born = factor(us_born, levels = c("Born in U.S.", "Born outside U.S.")),

    interview_lang = factor(interview_lang, levels = c("English", "Spanish")),

    # health status
    gen_health = factor(gen_health, levels = c("Excellent", "Very good",
                                               "Good", "Fair", "Poor")),
    hypertension = factor(hypertension, levels = c("No", "Yes")),

    # behaviors
    smoked_100   = factor(smoked_100,   levels = c("No", "Yes")),
    alcohol_ever = factor(alcohol_ever, levels = c("No", "Yes"))
  )
```

# 3. Restrict to the exam sample and check missingness

```{r mec-weights}
# drop people with zero MEC weight: respondents who did the interview but did
# not come in for the exam. We need the exam weight because BMI was measured
# in the exam.
df_analytic <- df %>% filter(WTMECPRP > 0)

cat("sample size after dropping zero MEC weights:", nrow(df_analytic), "\n")
cat("insured:",   sum(df_analytic$insured == "Insured"),   "\n")
cat("uninsured:", sum(df_analytic$insured == "Uninsured"), "\n")
```

```{r missing-check}
missing_check <- df_analytic %>%
  summarise(across(c(age, sex, race_eth, education, marital, us_born,
                     interview_lang, pir, bmi, gen_health,
                     hypertension, smoked_100, alcohol_ever),
                   ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(n_missing / nrow(df_analytic) * 100, 1)) %>%
  arrange(desc(n_missing))

print(missing_check)
# pir and alcohol_ever have the most missing (~14% and ~6.5%).
# complete-case analysis is used - svyglm drops missing rows automatically.
```

# 4. Survey design

```{r design}
# this is what makes NHANES estimates nationally representative
# id = the PSU (cluster), strata = the sampling stratum, weights = exam weight
nhanes_design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~WTMECPRP,
  data    = df_analytic,
  nest    = TRUE
)
```

```{r crude-rates}
# weighted crude screening rates before modeling
crude_rates <- svyby(~screened_bin, ~insured, nhanes_design, svymean, na.rm = TRUE)
crude_rates$pct <- round(crude_rates$screened_bin * 100, 1)
cat("weighted screening rates by insurance status:\n")
print(crude_rates[, c("insured", "pct")])
```

# 5. Models

## Model 1 — unadjusted

Just insurance alone, no covariates: the raw association before accounting for
anything else.

```{r m1}
m1 <- svyglm(
  screened_bin ~ insured,
  design = nhanes_design,
  family = quasibinomial(link = "logit")
)

summary(m1)

# exponentiate to get ORs, use df=Inf for CIs (see note under model 3)
m1_results <- as.data.frame(exp(cbind(OR = coef(m1), confint(m1, df = Inf))))
m1_results$p_value <- round(summary(m1)$coefficients[, "Pr(>|t|)"], 4)
cat("\nodds ratios:\n")
print(round(m1_results, 3))
```

## Model 2 — demographically adjusted

Adds age, sex, race/ethnicity, education, income, marital status, nativity, and
interview language as confounders — the core variables that differ between
insured/uninsured and also affect whether someone gets screened.

```{r m2}
m2 <- svyglm(
  screened_bin ~ insured + age + sex + race_eth +
    education + pir + marital + us_born + interview_lang,
  design = nhanes_design,
  family = quasibinomial(link = "logit")
)

summary(m2)

m2_results <- as.data.frame(exp(cbind(OR = coef(m2), confint(m2, df = Inf))))
m2_results$p_value <- round(summary(m2)$coefficients[, "Pr(>|t|)"], 4)
cat("\nodds ratios:\n")
print(round(m2_results, 3))
```

## Model 3 — fully adjusted

Adds BMI, general health, hypertension, smoking, and alcohol on top of model 2 —
health status and behavior that might also confound the relationship.

> **Note on CIs.** The masked NHANES PSU structure leaves very few residual
> degrees of freedom once this many variables are in the model, so the default
> CIs come out as 0 to Inf. We use `df = Inf` as a workaround (normal
> approximation), but the model-3 CIs are still unreliable. The ORs and
> p-values remain valid.

```{r m3}
m3 <- svyglm(
  screened_bin ~ insured + age + sex + race_eth +
    education + pir + marital + us_born + interview_lang +
    bmi + gen_health + hypertension + smoked_100 + alcohol_ever,
  design = nhanes_design,
  family = quasibinomial(link = "logit")
)

summary(m3)

m3_results <- as.data.frame(exp(cbind(OR = coef(m3), confint(m3, df = Inf))))
m3_results$p_value <- round(summary(m3)$coefficients[, "Pr(>|t|)"], 4)
cat("\nodds ratios:\n")
print(round(m3_results, 3))
```

# 6. Insurance effect across models

The main thing we care about: does the insurance OR stay consistent as we add
adjustment?

```{r summary-table}
get_insurance_row <- function(model, label) {
  or_ci <- exp(cbind(OR = coef(model), confint(model, df = Inf)))["insuredInsured", ]
  pval  <- summary(model)$coefficients["insuredInsured", "Pr(>|t|)"]
  data.frame(model    = label,
             OR       = round(or_ci["OR"], 3),
             CI_lower = round(or_ci["2.5 %"], 3),
             CI_upper = round(or_ci["97.5 %"], 3),
             p_value  = round(pval, 4))
}

insurance_summary <- rbind(
  get_insurance_row(m1, "unadjusted"),
  get_insurance_row(m2, "demographically adjusted"),
  get_insurance_row(m3, "fully adjusted")
)
rownames(insurance_summary) <- NULL
print(insurance_summary)

write.csv(insurance_summary, "insurance_effect_by_model.csv", row.names = FALSE)
cat("\nsaved to insurance_effect_by_model.csv\n")
```

```{r wald-tests}
# formal test of whether insurance adds anything beyond chance in each model
cat("model 1:\n"); print(regTermTest(m1, ~insured))
cat("model 2:\n"); print(regTermTest(m2, ~insured))
cat("model 3:\n"); print(regTermTest(m3, ~insured))
```

# 7. Session info

```{r session}
sessionInfo()
```
